Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Minor formatting fixes

The Third Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
held on 28-29 September 2015 in Boulder, Colorado, USA. Previous events in the
WSSSPE series are
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized in July 2014 jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}},
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14, and
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized in July 2015 again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.

@@

The workshop included multiple mechanisms for participation and encouraged team
building around solutions. It strongly encouraged participation of early-career
scientists, postdoctoral researchers, and graduate students, with funds provided
to the conference organizers by the Moore Foundation, the National Science
Foundation (NSF), and the Software Sustainability Institute, to support the
travel of potential participants who would not otherwise be able to attend,
younger participants and those from underrepresented groups. These funds allowed
16 additional participants to attend.

This report is based on collaborative notes taken with during the workshop,
which were linked from the GitHub issues that represented the potential and
actual working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}); the keynotes (\S\ref{sec:keynote}); a series of
lightning talks (\S\ref{sec:lightning}), intended to give an opportunity for
attendees to quickly highlight an important issue or a potential solution. The
report also gives summaries of action plans proposed by the eleven working
groups (\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}).
Finally, the report includes a list of registered attendees
(Appendix~\ref{sec:attendees}) and longer descriptions of the activities that
occurred in each the working groups
(Appendices~\ref{sec:appendix_best_practices}-\ref{sec:appendix_user_community}.)

@@

The first call for participation requested lightning talks, where authors could
make a brief statement about work that either had been done or was needed, with
the goal of contributing to the discussion of one or more working groups. There
was 24 lightning talks submitted, and after a peer-review process, 16 of these
were accepted, as discussed further in Section~\ref{sec:lightning}.

@@

WSSSPE3 broke free of the traditional Supercomputing Conference environment this
year, and in Turk's words, it spoke to the fact that scientific software comes
from many different types of inquiries, deployment, strategies for maintenance,
users; and ways of measuring the value of a piece of software. It appeared to
Turk that the supercomputing conferences generally adopt some top-down
approaches, whereas SciPy more often than not uses more bottom-up systems. The
essential messages perceived were also often bipolar: software getting harder in
supercomputing communities with exascale computing and optimization issues in
mind, but software becoming better in the SciPy group with emerging tools such
as Jupiter and productivity packages for research workflow. Admitting such
comparisons are somewhat unfair generalizations, Turk reminded the audience that
the different approaches bring different types of ideas to the table, but he
welcomed WSSSPE3 being conducted outside existing preconceptions.

Returning to the topic of his talk, Turk invited the audience to picture
scientific software as a flower on a landscape under the Sun, which may
represent in this analogy a number of measurable factors such as number of
citations; growth of a community and number of contributors; amount of funding;
prestigious prizes awarded; stability of the community in terms of leadership
transitions, serving community needs, not breaking test suites, and performance
on new architectures. But all these metrics are strictly speaking
\emph{proxies} for the values and the impact scientific software bears. What we
can measure does not give us direct insight---it just gives us proxies of
insight.

>>>>

The Third Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
WSSSPE series are
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized in July 2014 jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}},
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14, and
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized in July 2015 again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.

@@

The workshop included multiple mechanisms for participation and encouraged team
building around solutions. It strongly encouraged participation of early-career
scientists, postdoctoral researchers, and graduate students, with funds provided
to the conference organizers by the Moore Foundation, the National Science
Foundation (NSF), and the Software Sustainability Institute, to support the
travel of potential participants who would not otherwise be able to attend,
younger participants, and those from underrepresented groups. These funds allowed
16 additional participants to attend.

This report is based on collaborative notes taken with during the workshop,
which were linked from the GitHub issues that represented the potential and
actual working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}); the keynotes (\S\ref{sec:keynote}); a series of
lightning talks (\S\ref{sec:lightning}), intended to give an opportunity for
attendees to quickly highlight an important issue or a potential solution. The
report also gives summaries of action plans proposed by the eleven working
groups (\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}).
Finally, the report includes a list of registered attendees
(Appendix~\ref{sec:attendees}) and longer descriptions of the activities that
occurred in each the working groups
(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}.)

@@

The first call for participation requested lightning talks, where authors could
make a brief statement about work that either had been done or was needed, with
the goal of contributing to the discussion of one or more working groups. There
was 24 lightning talks submitted, and, after a peer-review process, 16 of these
were accepted, as discussed further in Section~\ref{sec:lightning}.

@@

WSSSPE3 broke free of the traditional Supercomputing Conference
environment this year, and in Turk's words, this spoke to the fact that scientific
software comes from many different types of inquiries, deployment, strategies
for maintenance, users; and ways of measuring the value of a piece of software.
It appeared to Turk that the supercomputing conferences generally adopt some
top-down approaches, whereas SciPy generally uses more bottom-up systems.
The essential messages
perceived were also often circumpolar: \katznote{this means ``situated around or inhabiting one of the earth's poles'' - not sure this is the right word} software getting harder in supercomputing
communities with exascale computing and optimization issues in mind, but software becoming better in the
SciPy group with emerging tools such as Jupiter and productivity packages for research workflow.
Admitting such comparisons are somewhat unfair generalizations, Turk reminded the
audience that the different approaches bring different types of ideas to the
table, but he welcomed WSSSPE3 being conducted outside existing preconceptions.
